Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging

### 2023-01-06 script/manuscript_MEE.Rmd
added reporting for matching morpho to molecular

```{r Required Libraries, message=F, warning=F, results='hide', echo = F}
library(here)
library(tidyverse)
library(bookdown)
library(knitr)
suppressWarnings(library(kableExtra))
pNumC <- function(...) prettyNum(...,big.mark = ',')
#print(citation('spdep'), bibtex=TRUE)
```

@@

```{r, echo = F}
a <- c('Trifolium', 'Lupinus', 'Glyccrrhiza', 'Mitella', 'Geum')
b <- c('Lupinus', 'Lathyrus', 'Potentilla', 'Androsace', 'Bistorta', 'Vicia')
c <- c('Jeffersonia', 'Micranthes', 'Prunus', 'Delphinium', 'Androsace', 
       'Penstemon', 'Orthocarpus', 'Scuttelaria', 'Aquilegia', 'Castilleja', 'Draba')
d <- c('Salix', 'Boechera')

@@

corb <- read.csv('../data/Corbiculae_collections.csv', header = F) %>% 
  mutate(V1 = str_remove(V1, 'dark|orange')) %>% 
  filter(!str_detect(V1, 'bifarius')) %>% 
  separate(V1, into = c('sample.id', 'genus', 'species', 'caste',
                        'plant.genus', 'plant.species')) %>% 
  filter(plant.genus == 'Mertensia')

pmorph <- read.csv('../data/morpho_pollen.csv')
mertensia_samples <- pmorph %>% 
  group_by(sample.id) %>% 
  filter(morphotype == 'MERTENSIA') %>% 
  select(sample.id, sum)

# Do we detect mertensia, asteraceae 1, asteraceae 2, in all molecular samples which they are present in the pollen samples of?

blst <- read.csv('../data/classified_blast.csv')

@@

@@

blst <- read.csv('../data/classified_blast.csv') %>% 
  distinct(sample.id)

blst <- read.csv('../data/classified_blast.csv') %>% 
  mutate(across(where(is.character), ~ replace_na(., 'UNKNOWN'))) %>% 
  arrange(sample.id, -Prcnt_seqs) %>% 
  inner_join(., corb, by = 'sample.id')

>>>>

```{r Required Libraries, message=F, warning=F, results='hide', echo = F}
library(here)
library(tidyverse)
library(bookdown)
library(knitr)
suppressWarnings(library(kableExtra))
pNumC <- function(...) prettyNum(...,big.mark = ',')

prcnt_detect <- function(x, Lprop, Hprop){
  if(missing(Lprop)){Lprop <- 0}

  y <- x[x$Prop > Lprop & x$Prop < Hprop,]; total <- nrow(y)
  detected <- round((nrow(y[y$DETECTED == 1,])/total) * 100, 1)
  Mdn <- round(median(y$Prop),3)

  return(writeLines(paste0('were detected in ',  detected, 
                '% of samples where the proportion of target grains were between ', 
                Lprop, '-', Hprop,  ' (n = ', total, ' Mdn = ', Mdn, ')')))
}
```

@@

```{r, echo = F}
#a <- c('Trifolium', 'Lupinus', 'Glyccrrhiza', 'Mitella', 'Geum')
#b <- c('Lupinus', 'Lathyrus', 'Potentilla', 'Androsace', 'Bistorta', 'Vicia')
#c <- c('Jeffersonia', 'Micranthes', 'Prunus', 'Delphinium', 'Androsace', 
#       'Penstemon', 'Orthocarpus', 'Scuttelaria', 'Aquilegia', 'Castilleja', 'Draba')
#d <- c('Salix', 'Boechera')

@@

pmorph <- read.csv('../data/morpho_pollen.csv') 
blst <- read.csv('../data/blast_seqs_classified.csv') 
blst_seqs <- distinct(blst, sample.id)

pmorph <- filter(pmorph, sample.id %in% blst$sample.id) %>% 
  group_by(sample.id) %>% 
  mutate(Prop = sum/sum(sum))

mertensia_samples <- pmorph %>% 
  filter(morphotype == 'MERTENSIA') %>% 
  select(sample.id, Prop, sum)

mertensia_samples <- filter(mertensia_samples, sample.id %in% blst_seqs$sample.id)

blst_mertensia <- blst %>% 
  filter(Genus == 'Mertensia') %>% 
  select(sample.id,  Prcnt_seqs) %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_seqs = sum(Prcnt_seqs)) %>% 
  distinct(sample.id, .keep_all = T)

# set up as evaluation table !?

mertensia_detected <- mertensia_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_mertensia$sample.id, 1, 0)) 
 # sample 26 - very strange !!!! this needs investigation

M1 <- glm(DETECTED ~ Prop, mertensia_detected, family = binomial)
M.df <- data.frame(Prop = seq(0, 1, 0.0001))
M.df$DETECTED <- predict(M1, newdata=M.df, type="response")

ggplot(mertensia_detected, aes(x=Prop, y=DETECTED)) +
  stat_smooth(method="glm", se=T, fullrange=TRUE, 
              method.args = list(family=binomial))  +
  theme_bw()

# Repeat for Asteraceae 1 & 2

blst_aster <- blst %>% 
  filter(Family == 'Asteraceae') %>% 
  select(sample.id,  Prcnt_seqs, Genus) %>% 
  group_by(sample.id) %>% 
  distinct(sample.id, .keep_all = T)

# ASTER 2 HERE...
aster2_samples <- pmorph %>% 
  filter(morphotype == 'ASTERACEAE_2')

blst_aster2 <- filter(blst_aster, Genus %in% 
                            c('Taraxacum', 'Helianthella', 'Scabrethia'))

aster2_detected <- aster2_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_aster2$sample.id, 1, 0))

M1 <- glm(DETECTED ~ Prop, aster2_detected, family = binomial)
M.df <- data.frame(Prop = seq(0, 1, 0.0001))
M.df$DETECTED <- predict(M1, newdata=M.df, type="response")

ggplot(aster2_detected, aes(x=Prop, y=DETECTED)) +
  stat_smooth(method="glm", se=T, fullrange=TRUE, method.args = list(family='binomial'))  +
  theme_bw()

# ASTER 1 HERE... ? WERE ALL OF THESE SPECIES PRUNED... ?
aster1_samples <- pmorph %>% 
  filter(morphotype == 'ASTERACEAE_1')

blst_aster1 <- filter(blst_aster, !Genus %in% c('Taraxacum', 'Helianthella', 'Scabrethia') )

aster1_detected <- aster1_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_aster1$sample.id, 1, 0))

ggplot(aster1_detected, aes(x=Prop, y=DETECTED)) +
  stat_smooth(method="glm", se=T, fullrange=TRUE, method.args = list(family='binomial'))  +
  theme_bw()

# ALL HIGH NUMBER ASTERACEAE

aster_samples <- pmorph %>% 
  filter(str_detect(morphotype, 'ASTER.*'))
aster_detected <- aster_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_aster$sample.id, 1, 0))

rm(M1, M.df, M.df,  mertensia_samples, blst, blst_aster, 
   blst_aster1, blst_aster2, aster1_samples,  aster2_samples, 
   a, b, c, d, blast_mertensia, aster_samples)
```

The pollen samples contained three morphotypes which could readily be identified. Two of these mapped to monophyletic clades (Boraginaceae & Heliantheae Alliance), and one to a paraphyletic clade (Asteraceae less Heliantheae).
Boraginaceae grains `r prcnt_detect(mertensia_detected, Lprop = 0.01, Hprop = 1)`.  Asteraceae type 1, non-helianthoids, `r prcnt_detect(aster1_detected, Lprop = 0.001, Hprop = 0.01)` Asteraceae type 2, Helianthoids, `r prcnt_detect(aster2_detected, Lprop = 0.001, Hprop = 0.01)`. Both morphotypes of Asteraceae pollen `r prcnt_detect(aster_detected, Lprop = 0.01, Hprop = 1)`.

@@

```{r}
rm(mertensia_detected, aster1_detected, aster2_detected, aster_detected)
```

@@

blst <- read.csv('../data/blast_seqs_classified.csv') %>% 
  distinct(sample.id)

blst <- read.csv('../data/blast_seqs_classified.csv') %>% 
  mutate(across(where(is.character), ~ replace_na(., 'UNKNOWN'))) %>% 
  arrange(sample.id, -Prcnt_seqs) %>% 
  inner_join(., corb, by = 'sample.id')
